\textit{k}-mer-based GWAS enhances the discovery of causal variants and candidate genes in soybean

### 2023-07-20 additional_files/additional_file_1.tex
Add a figure about sequencing depth and some material in the discussion.
Version submitted as revision

@@

>>>>

% A table with the number of k-mers foundd by each analysis
\begin{table}
	\centering
	\caption[Number of k-mers per GWAS analysis]{Number of significant \textit{k}-mers identified per GWAS analysis}
	\label{kmer_count_table}

	\begin{threeparttable}\footnotesize

			\csvreader[head to column names,
			tabular = lrrr@{\hskip 3pt}l,
			table head = \toprule GWAS analysis\tnote{a} & Raw\tnote{b} & Filtered\tnote{c} & At known loci & (\%)\tnote{d} \\ \midrule,
			table foot=\bottomrule,
			before reading={\catcode`\#=12},
			after reading={\catcode`\#=6},
			/csv/separator=semicolon]%
			{../tables/kmer_count_table.csv}{}{\gwas & \raw & \filtered & \nsignals & (\percent)}

				\begin{tablenotes}
				\item[a] GWAS numbers refer to numbers in tables \ref{FLOWER.COLOR_gwas_table}
					to \ref{MATURITY.GROUP_gwas_table}.
				\item[b] Number of significant \textit{k}-mers in raw GWAS output.
				\item[c] Number of significant \textit{k}-mers after filtering for
					mapping quality, minimum count and redundancy.
				\item[d] Number of \textit{k}-mers overlapping known loci and
					percentage relative to number of filtered \textit{k}-mers.
				\end{tablenotes}

	\end{threeparttable}
\end{table}

@@

% Figure for the proportion of genome covered by only one read as a function of sequencing depth
\begin{cfigure}
	\includegraphics[width = 6.5in]{seqdepth}
	\caption[Proportion of the genome covered by one read versus sequencing depth]{%
		Proportion of the genome covered by only one read as a function
	of the sequencing depth of 363 Illumina whole-genome sequencing soybean
	samples used for GWAS in this study.}
	\label{seqdepth-figure}
\end{cfigure}

### 2023-07-20 additional_files/manuscript.tex
Version submitted as revision

Moreover, we pruned the dataset using the \texttt{-\-indep-pairwise} option of
PLINK v. 1.90b5.3 \shortcite{purcell2007} with a window of 1,000 markers, a
step of 100 markers and an $r^2$ LD threshold of 0.9. This pruning step was
implemented to reduce the computational requirements of the GWAS analysis by
removing co-segregating markers. Markers located on unanchored scaffolds were
also removed. The resulting dataset was used for GWAS and comprised 773,060
SNPs and 151,570 indels. Pruned markers were still used for GWAS at targeted
loci when filtered markers identified significant associations (more details
in the section describing the analysis of significant signals).

@@

We used the approach outlined by \shortciteA{voichek2020} for \emph{k}-mer-based GWAS
and their recommendations on the
associated GitHub page (v. 0.2-beta, \url{https://github.com/voichek/kmersGWAS}) to generate
a presence/absence \emph{k}-mer table for use in GWAS.  The first step of
the approach involved counting all \emph{k}-mers of length 31 (31-mers) present
in the trimmed (FASTQ) reads for each accession using KMC3 v. 3.2.1
\shortcite{kokot2017}. We selected $k = 31$ based on demonstrated performance in previously
published studies \shortcite{rahman2018, voichek2020, he2021}.
\emph{k}-mers were counted twice, the first time using their
canonized form (the first of either the observed \textit{k}-mer or its reverse
complement in lexicographical order) and the second time using the observed
\emph{k}-mers themselves (non-canonized form). The two sets of \textit{k}-mers
were then combined and filtered by
keeping only those seen in at least 5 accessions, and observed in both canonized
and non-canonized form in at least 20\% of the accessions in which they are found.
Only \textit{k}-mers that occurred twice in a given sample were considered to
reduce the impact of sequencing errors.
We then generated a table indicating the presence or absence of each
\emph{k}-mer in all accessions. This table was used to compute a kinship
table for use in GWAS and was also used directly as input genotypes for the
GWAS analyses. We used all \emph{k}-mers with a MAF $\geq$ 0.02 as input genotypes
for both the computation of kinship and GWAS analysis. The total number of \textit{k}-mers
in this filtered table was 637,605,529.

@@

Since two or three different GWAS analyses were conducted for some of the
traits, a total of 22 GWAS analyses were performed using each of the three
approaches.  Given the large volume of data that this represents, we focus here
on the most noteworthy results. Readers interested in results at loci that are
not discussed here are referred to Supplemental file 1. In particular, Supplemental Table
S11 provides a summary of the
\textit{p}-values obtained by each approach at each of the loci considered in
this study.

@@

In addition to verifying known loci, we also observed potentially novel loci in
the \textit{k}-mer GWAS analysis of stem termination and pod color.  When
comparing accessions with indeterminate and semi-determinate stem termination
types, we observed previously undocumented signals on chromosomes Gm11, Gm16
and Gm18 in addition to the known \textit{Dt1} locus (Figure
\ref{stem-termination-main-figure}a).  Analysis of the pairwise LD between
significant \textit{k}-mers found no LD between these signals, suggesting that
they may represent \textit{bona fide} loci associated with stem termination
type (Figure \ref{stem-termination-main-figure}b). Similarly, a \textit{k}-mer
GWAS analysis comparing accessions with black and brown pods found a previously
undocumented signal on chromosome Gm15 (Supplemental Figure S43).
\textit{k}-mers at this new locus
showed no LD with \textit{k}-mers at the known \textit{L1} locus and may
therefore represent a distinct locus (Supplemental Figure 
S44). The observed signal overlaps a total of 27 genes
(Supplemental Figure S45), but none of those are annotated
as having an obvious role in controlling pod color.

@@

While we have have shown that using \emph{k}-mers for GWAS in soybean performed
well compared to other GWAS methods, this approach does present some unique
challenges and limitations. One of these challenges is the
interpretation of \textit{k}-mer-based GWAS given the greater possibility
of false positive associations than in conventional GWAS approaches. In this
study, we noticed apparently spurious associations with \textit{k}-mers
scattered throughout the genome were common. This was especially the case for
highly significant loci whose signal spanned several megabases such as \textit{W}
and \textit{T}. A look at the Q-Q plots derived from \textit{k}-mer-based GWAS
showed that these usually had the same shape as the Q-Q plots derived from SNP-
and SV-based GWAS, suggesting that population structure was properly controlled.
We believe that a more likely culprit for these spurious associations is the
difficulty in linking \textit{k}-mers to genomic coordinates. At loci that
span a large interval, it is indeed be highly likely that some of the \textit{k}-mers
coming from the region map elsewhere in the genome. This interpretation is supported
by the fact that the LD between presumably spurious \textit{k}-mers and
\textit{k}-mers from target loci was generally high (e.g. Figures S5 and S47).
Overall, the risk of false association due to population structure and the identification
of an appropriate significance threshold seems well controlled, but work remains
to be done in reducing the occurrence of spurious associations.

There are obvious limitations to the relying on the presence/absence of
\emph{k}-mers for GWAS. For example, this approach may not be appropriate in
species where heterozygosity is common (i.e., most species other than inbred
crops) as the \emph{k}-mers associated with the presence of both alleles will
be observed at heterozygous loci. Moreover, although the presence/absence of
\emph{k}-mers can detect variation independently of its type, some types of
variation may be more difficult to detect than others.  For example, copy
number variation at a locus may not be detected from presence/absence alone if
all copies are identical and occur in tandem, such that no \textit{k}-mer can
differentiate between two copies and more. For similar reasons, \emph{k}-mers
may also fail to capture variation occurring in repeated sequences. Alternative
models based on \emph{k}-mer counts instead of presence/absence have been
developed \shortcite{rahman2018, he2021} which may solve issues like these and
broaden the scope of variants that can be identified from \emph{k}-mers.
Nevertheless, \shortciteA{voichek2020} demonstrated that 98.9\% of known
translocations and inversions as well as 99.6\% of known insertions/deletions
in \textit{Arabidopsis thaliana} could be tagged by unique 31-mers. It is
therefore likely that the vast majority of variants can be detected through the
presence/absence of \textit{k}-mers.

In addition to the link between \textit{k}-mer presence/absence and sequence
variation, one may wonder to what extent errors may occur in declaring the
presence or absence of a \textit{k}-mer based on Illumina WGS data. If we
neglect the possibility of contamination, the most likely error would be to
declare a \textit{k}-mer as absent simply because sequencing depth was
insufficient. In this study, we declared \textit{k}-mers as present
based on a minimum of two occurrences. Therefore, a \textit{k}-mer may be
wrongly declared as absent if it is observed only once or not at all.  To
quantify this error, we compared the number of times that a given position was
observed only once in a sample to the average sequencing depth of that sample.
We found that most samples with a sequencing depth between 10 and 20X had
a proportion of only 0.5 to 1.5\% of positions covered by only one read.
Therefore, the risk of falsely declaring a \textit{k}-mer as absent from
our dataset was low, and in any case much lower than typical error rates
involved in calling and genoyping SVs \shortcite<e.g.>[]{kosugi2019, ebler2022}.
The robustness of \textit{k}-mer presence/absence to sequencing depth is also
demonstrated by \shortciteA{gaurav2022}.

>>>>

Moreover, we pruned the dataset using the \texttt{-\-indep-pairwise} option of
PLINK v. 1.90b5.3 \shortcite{purcell2007} with a window of 1,000 markers, a
step of 100 markers and an $r^2$ LD threshold of 0.9. This pruning step was
implemented to reduce the computational requirements of the GWAS analysis by
removing co-segregating markers. Markers located on unanchored scaffolds were
also removed. The resulting dataset was used for GWAS and comprised 773,060
SNPs and 151,570 indels. Pruned markers were still used for GWAS at targeted
loci where filtered markers identified significant associations (more details
in the section describing the analysis of significant signals).

@@

We used the approach outlined by \shortciteA{voichek2020} for \emph{k}-mer-based GWAS
and their recommendations on the
associated GitHub page (v. 0.2-beta, \url{https://github.com/voichek/kmersGWAS}) to generate
a presence/absence \emph{k}-mer table for use in GWAS.  The first step of
the approach involved counting all \emph{k}-mers of length 31 (31-mers) present
in the trimmed (FASTQ) reads for each accession using KMC3 v. 3.2.1
\shortcite{kokot2017}. We selected $k = 31$ based on demonstrated performance in previously
published studies \shortcite{rahman2018, voichek2020, he2021}.
\emph{k}-mers were counted twice, the first time using their
canonized form (the first of either the observed \textit{k}-mer or its reverse
complement in lexicographical order) and the second time using the observed
\emph{k}-mers themselves (non-canonized form). The two sets of \textit{k}-mers
were then combined and filtered by
keeping only those seen in at least 5 accessions, and observed in both canonized
and non-canonized form in at least 20\% of the accessions in which they are found.
To reduce the potential impact of sequencing errors, only \textit{k}-mers that
occurred at least twice in a given sample were considered.
We then generated a table indicating the presence or absence of each
\emph{k}-mer in all accessions. This table was used to compute a kinship
table for use in GWAS and was also used directly as input genotypes for the
GWAS analyses. We used all \emph{k}-mers with a MAF $\geq$ 0.02 as input genotypes
for both the computation of kinship and GWAS analysis. The total number of \textit{k}-mers
in this filtered table was 637,605,529.

@@

Since two or three different GWAS analyses were conducted for some of the
traits, a total of 22 GWAS analyses were performed using each of the three
approaches.  Given the large volume of data that this represents, we focus here
on the most noteworthy results. Readers interested in results at loci that are
not discussed here are referred to Supplemental file 1. In particular, Supplemental Table
S11 provides a summary of the
\textit{p}-values obtained by each approach at each of the loci considered in
this study. Supplemental Table S12 also shows some results specific to \textit{k}-mer-based
analyses.

@@

In addition to verifying known loci, we also observed potentially novel loci in
the \textit{k}-mer GWAS analysis of stem termination and pod color.  When
comparing accessions with indeterminate and semi-determinate stem termination
types, we observed previously undocumented signals on chromosomes Gm11, Gm16
and Gm18 in addition to the known \textit{Dt1} locus (Figure
\ref{stem-termination-main-figure}a).  Analysis of the pairwise LD between
significant \textit{k}-mers found no LD between these signals, suggesting that
they may represent \textit{bona fide} loci associated with stem termination
type (Figure \ref{stem-termination-main-figure}b). We note that the signal newly
identified on Gm18 is located 5 Mb away from the already documented \textit{Dt2}
locus and may therefore correspond to that locus.
Similarly, a \textit{k}-mer
GWAS analysis comparing accessions with black and brown pods found a previously
undocumented signal on chromosome Gm15 (Supplemental Figure S43).
\textit{k}-mers at this new locus
showed no LD with \textit{k}-mers at the known \textit{L1} locus and may
therefore represent a distinct locus (Supplemental Figure 
S44). The observed signal overlaps a total of 27 genes
(Supplemental Figure S45), but none of those are annotated
as having an obvious role in controlling pod color.

@@

While we have have shown that using \emph{k}-mers for GWAS in soybean performed
well compared to other GWAS methods, this approach does present some unique
challenges and limitations. One of these challenges is the
interpretation of \textit{k}-mer-based GWAS given the greater possibility
of false positive associations than in conventional GWAS approaches. In this
study, we noticed that spurious associations, appearing as \textit{k}-mers
scattered throughout the genome, were common. This was especially the case for
traits with highly significant signals that spanned several megabases, such as flower color
and pubescence color (Supplemental Table S12). Q-Q plots derived from \textit{k}-mer-based GWAS
generally displayed the same shape as the Q-Q plots derived from SNP-
and SV-based GWAS, suggesting that population structure was properly controlled.
Despite the very large number of tests conducted, the approach developed by
\shortciteA{voichek2020} for setting significance thresholds and controlling
for population structure therefore appears appropriate.
We believe that a more likely culprit for these spurious associations is the
difficulty in linking \textit{k}-mers to genomic coordinates. At loci that
span a large interval, it is indeed highly likely that some of the \textit{k}-mers
coming from the target region will map elsewhere in the genome. This interpretation is supported
by the fact that the LD between presumably spurious \textit{k}-mers and
\textit{k}-mers from target loci was generally high (e.g. Supplemental Figures S5 and S47).
Until further work resolves these issues, we recommend users of
\textit{k}-mer-based GWAS to assess their results critically and to be aware
that spurious associations are likely to occur.

There are obvious limitations to relying on the presence/absence of
\emph{k}-mers for GWAS. For example, this approach may not be appropriate in
species where heterozygosity is common (i.e., most species other than inbred
crops) as the \emph{k}-mers associated with the presence of both alleles will
be observed at heterozygous loci. Moreover, although the presence/absence of
\emph{k}-mers can detect variation independently of its type, some types of
variation may be more difficult to detect than others.  For example, copy
number variation at a locus may not be detected from presence/absence alone if
all copies are identical and occur in tandem, such that no \textit{k}-mer can
differentiate between two copies and more. For similar reasons, \emph{k}-mers
may also fail to capture variation occurring in repeated sequences. Alternative
models based on \emph{k}-mer counts instead of presence/absence have been
developed \shortcite{rahman2018, he2021} which may solve issues like these and
broaden the scope of variants that can be identified from \emph{k}-mers.
Nevertheless, \shortciteA{voichek2020} demonstrated that 98.9\% of known
translocations and inversions as well as 99.6\% of known insertions/deletions
in \textit{Arabidopsis thaliana} could be tagged by unique 31-mers. It is
therefore likely that the vast majority of variants can be detected through the
presence/absence of \textit{k}-mers.

A related problem is the extent to which errors may occur in declaring the
presence or absence of a \textit{k}-mer based on Illumina WGS data. If we
neglect the possibility of contamination, the most likely error would be to
declare a \textit{k}-mer as absent simply because sequencing depth was
insufficient. In this study, we declared \textit{k}-mers as present
based on a minimum of two occurrences. Therefore, a \textit{k}-mer may be
wrongly declared as absent if it is observed only once or not at all.  To
quantify this error, we compared the number of positions that were
observed only once in a sample to the average sequencing depth of that sample.
We found that most samples with a sequencing depth between 10 and 20X had
between 0.5 to 1.5\% of positions covered by a single read (Supplemental Figure S46).
Therefore, the risk of falsely declaring a \textit{k}-mer as absent from
our dataset was low, and in any case much lower than typical error rates
involved in calling and genoyping SVs \shortcite<e.g.>[]{kosugi2019, ebler2022}.
The robustness to low sequencing depth of GWAS analyses based on \textit{k}-mer
presence/absence was also demonstrated by \shortciteA{gaurav2022}.
